### 2021-04-19 paper/maxdet.tex
Added the first theorem for the semidefinite case

\section{Independence, maximal determinant, and entropy}
\subsection{A simpler problem}
Before we answer the question of the last section in full generality we look at a simpler case to determine how we want to complete the matrix $M$. As in the previous section we denote by $I$ be the set of stochastic processes. For this simple case we assume that there are two sets $\alpha, \beta\subset I$ with $\alpha\cup\beta = I$ and $\alpha\cap\beta = \gamma\ne \emptyset$. For any $\delta\subset I$ we will denote by
\begin{equation}
	M[\delta]
\end{equation}
the principal submatrix of $M$ with row and column indices in $\delta$. For two sets $\delta,\eta\subset I$ we denote by 
\begin{equation}
	M[\delta,\eta]
\end{equation}
the submatrix of $M$ with row indices in $\delta$ and column indices in $\eta$. We assume that
\begin{equation}
	M[\alpha] \text{\ and\ } M[\beta] 
\end{equation}
are known. The matrix $M$ can then be written in the form
\begin{equation}\label{eqn.corrmatrix}
	M = \begin{pmatrix}
		C & D & W \\
		D^T & E & F \\
		W^T & F^T & G
	\end{pmatrix},
\end{equation}
where we make the following identifications:
\begin{align}
	M[\alpha] & = \begin{pmatrix}
		C & D  \\
		D^T & E 
	\end{pmatrix} \\
	M[\beta] & = \begin{pmatrix}
		E & F  \\
		F^T & G 
	\end{pmatrix} \\
	M[\gamma] & = E \\
	M[ \alpha-\gamma, \alpha-\gamma] & = C \\
	M[ \alpha-\gamma, \gamma] & = D \\
	M[ \alpha-\gamma, \beta-\gamma] & = W \\
	M[ \gamma, \beta-\gamma] & = F \\
	M[ \beta-\gamma, \beta-\gamma] & = F
\end{align}
The part of the matrix that we need to determine is $W$. If the overlap $E$ didn't exist we would just set $W=0$ and thus ensure that $\alpha$ and $\beta$ are independent. This is the desirable result because any other choice would assume information about the correlation between $\alpha$ and $\beta$ that we do not have. Since the overlap $E$ is present, we have to make a different choice.

\subsection{Conditioning \& Schur complement}
For $\delta\subset I$ let us assume that $M[\delta]$ is invertible (we will weaken this assumption later). Then let 
\begin{equation}
	M/M[\delta] = M[I-\delta,I-\delta] - M[I-\delta, \delta]M[\delta]^{-1}M[\delta, I-\delta]
\end{equation}
be the Schur complement\cite{hayns} (see \cite{horn} for a an overview of the properties of the Schur complement) of $M[\delta]$ in $M$. In our context the Schur complement arises naturally when we ask for the correlation matrix of a subset of the stochastic variables in $I$ conditioned on the values of the remaining variables. If the variables that we are conditioning on are in $\gamma$ then $M/M[\gamma]$ is the correlation matrix for the variables in $I-\gamma$(for a proof see \cite[chapter 8]{rao} and \cite{cottle}).

@@

\subsection{Maximal determinant}
Another way to characterize the choice of $W$ in equation (\ref{eqn.choice}) is by looking at the determinant of $M$. In general we have 
\begin{align}
	\det(M) & = \det(M[\gamma]) \det(M/M[\gamma]) \\
	& \le \det(M[\gamma]) \det(M[\alpha]/M[\gamma]) \det(M[\beta/M[\gamma]),\label{eqn.upperbound}
\end{align}
(see \cite{horn} for the first equality), where we have used the Fischer inequality in the second line. With $W$ as in equation (\ref{eqn.choice}) the Schur complement $M/M[\gamma]$ becomes block-diagonal and we obtain
\begin{equation}
	\det(M) = \det(M[\gamma]) \det(M[\alpha]/M[\gamma]) \det(M[\beta/M[\gamma]).
\end{equation}
The determinant thus reaches the upper bound of equation (\ref{eqn.upperbound}). Choosing $W$ as in equation (\ref{eqn.choice}) we thus maximize the determinant of $M$.

\subsection{Zeroes \& the inverse matrix}
A remarkable formula due to Babachiewicz \cite{horn} expresses the inverse of a matrix in terms of the Schur complement. For a matrix
\begin{equation}
	\Sigma = \begin{pmatrix}
		\Sigma_{11} & \Sigma_{12} \\ 
		\Sigma_{21} & \Sigma_{22}
	\end{pmatrix}
\end{equation}
we have
\begin{equation}
	\Sigma^{-1} = \begin{pmatrix}
		\Sigma_{11}^{-1} + \Sigma_{11}^{-1}\Sigma_{12}(\Sigma/\Sigma_{11})^{-1}\Sigma_{21}\Sigma_{11}^{-1} &-\Sigma_{11}^{-1} \Sigma_{12} (\Sigma/\Sigma_{11})^{-1} \\ 
		-(\Sigma/\Sigma_{11})^{-1}\Sigma_{21}\Sigma_{11}^{-1} & (\Sigma/\Sigma_{11})^{-1}
	\end{pmatrix}.
\end{equation}
This formula has the remarkable property that only the inverses 
\begin{equation}
	\Sigma_{11}^{-1} \text{\ and\ } (\Sigma/\Sigma_{11})^{-1}
\end{equation}
appear in it. With our choice of $W$ in equation (\ref{eqn.choice}) the Schur complement $M/E$ is block diagonal. This means that also its inverse is block diagonal. Indeed, we have
\begin{equation}
	(M/E)^{-1} = \begin{pmatrix}
		(M[\alpha]/E)^{-1} & 0 \\
		0 & (M[\beta]/E)^{-1}
	\end{pmatrix}.
\end{equation}
This allows us to write the inverse of $M$ explicitly:
\begin{equation}
	M^{-1} = \begin{pmatrix}
		(M[\alpha]/E)^{-1} & -(M[\alpha]/E)^{-1} D E^{-1} & 0 \\
		- E^{-1} D^T (M[\alpha]/E)^{-1} & \Xi & - E^{-1} F (M[\beta]/E)^{-1} \\
		0 & - (M[\beta]/E)^{-1} F^T E^{-1} & (M[\beta]/E)^{-1}
	\end{pmatrix},
\end{equation}
with
\begin{equation}
	\Xi = E^{-1} + E^{-1}\left(D^T(M[\alpha]/E)^{-1}D + F (M[\beta]/E)^{-1} F^T\right)E^{-1}.
\end{equation}
We see that the inverse of $M$ has zeroes in exactly those places where the undetermined matrix $W$ used to be.

@@

\subsection{The semidefinite case}

@@

\begin{thebibliography}{MM}
	\bibitem{heston} Steven L. Heston, A Closed-Form Solution for Options with Stochastic Volatility with Applications to Bond and Currency Options, The Review of Financial Studies, \textbf{6}(2), April 1993, 327--343.
	\bibitem{hayns} Emilie Virginia Haynsworth, Determination of the inertia of a partitioned Hermitian matrix,  Linear  Algebra and its Applications \textbf{1}, 1968, 73--81. 
	\bibitem{horn} Roger A. Horn, Basic  Properties  of  the  Schur  Complement. In Fuzhen Zhang (Ed.) The Schur complement and its application (17--46). Springer, 2005.
	\bibitem{rao} C. Radharkishna Rao, Linear Statistical Inference and its Applications, 2nd edition, John Wiley \& Sons, Inc, 2002.
	\bibitem{cottle} Richard W. Cottle, Manifestations of the Schur complement, Linear Algebra and its Applications \textbf{8}, 1974, 189--211. 
\end{thebibliography}

>>>>

\section{Independence, maximal determinant, and entropy}
\subsection{A simpler problem}
Before we answer the question of the last section in full generality we look at a simpler case to determine how we want to complete the matrix $M$. As in the previous section we denote by $I$ be the set of stochastic processes. For this simple case we assume that there are two sets $\alpha, \beta\subset I$ with $\alpha\cup\beta = I$ and $\alpha\cap\beta = \gamma\ne \emptyset$. For any $\delta\subset I$ we will denote by
\begin{equation}
	M[\delta]
\end{equation}
the principal submatrix of $M$ with row and column indices in $\delta$. For two sets $\delta,\eta\subset I$ we denote by 
\begin{equation}
	M[\delta,\eta]
\end{equation}
the submatrix of $M$ with row indices in $\delta$ and column indices in $\eta$. We assume that
\begin{equation}
	M[\alpha] \text{\ and\ } M[\beta] 
\end{equation}
are known. The matrix $M$ can then be written in the form
\begin{equation}\label{eqn.corrmatrix}
	M = \begin{pmatrix}
		C & D & W \\
		D^T & E & F \\
		W^T & F^T & G
	\end{pmatrix},
\end{equation}
where we make the following identifications:
\begin{align}
	M[\alpha] & = \begin{pmatrix}
		C & D  \\
		D^T & E 
	\end{pmatrix} \\
	M[\beta] & = \begin{pmatrix}
		E & F  \\
		F^T & G 
	\end{pmatrix} \\
	M[\gamma] & = E \\
	M[ \alpha-\gamma, \alpha-\gamma] & = C \\
	M[ \alpha-\gamma, \gamma] & = D \\
	M[ \alpha-\gamma, \beta-\gamma] & = W \\
	M[ \gamma, \beta-\gamma] & = F \\
	M[ \beta-\gamma, \beta-\gamma] & = G
\end{align}
The part of the matrix that we need to determine is $W$. If the overlap $E$ didn't exist we would just set $W=0$ and thus ensure that $\alpha$ and $\beta$ are independent. This is the desirable result because any other choice would assume information about the correlation between $\alpha$ and $\beta$ that we do not have. Since the overlap $E$ is present, we have to make a different choice.

\subsection{Conditioning \& the Schur complement}
For $\delta\subset I$ let us assume that $M[\delta]$ is invertible (we will weaken this assumption later). Then let 
\begin{equation}
	M/M[\delta] = M[I-\delta,I-\delta] - M[I-\delta, \delta]M[\delta]^{-1}M[\delta, I-\delta]
\end{equation}
be the Schur complement\cite{hayns} (see \cite{horn} for a an overview of the properties of the Schur complement) of $M[\delta]$ in $M$. In our context the Schur complement arises naturally when we ask for the correlation matrix of a subset of the stochastic variables in $I$ conditioned on the values of the remaining variables. If the variables that we are conditioning on are in $\gamma$ then $M/M[\gamma]$ is the correlation matrix for the variables in $I-\gamma$(for a proof see \cite[chapter 8]{rao} and \cite{cottle}).

@@

\subsection{Maximal determinant}
Another way to characterize the choice of $W$ in equation (\ref{eqn.choice}) is by looking at the determinant of $M$. In general we have 
\begin{align}
	\det(M) & = \det(M[\gamma]) \det(M/M[\gamma]) \\
	& \le \det(M[\gamma]) \det(M[\alpha]/M[\gamma]) \det(M[\beta/M[\gamma]),\label{eqn.upperbound}
\end{align}
(see \cite{horn} for the first equality), where we have used the Fischer inequality in the second line. With $W$ as in equation (\ref{eqn.choice}) the Schur complement $M/M[\gamma]$ becomes block-diagonal and we obtain
\begin{equation}
	\det(M) = \det(M[\gamma]) \det(M[\alpha]/M[\gamma]) \det(M[\beta/M[\gamma]).
\end{equation}
The determinant thus reaches the upper bound of equation (\ref{eqn.upperbound}). Choosing $W$ as in equation (\ref{eqn.choice}) thus maximizes the determinant of $M$.

\subsection{Zeroes \& the inverse matrix}
A remarkable formula due to Babachiewicz \cite{horn} expresses the inverse of a matrix in terms of the Schur complement. For a matrix
\begin{equation}\label{eqn.sigma}
	\Sigma = \begin{pmatrix}
		\Sigma_{11} & \Sigma_{12} \\ 
		\Sigma_{21} & \Sigma_{22}
	\end{pmatrix}
\end{equation}
we have
\begin{equation}
	\Sigma^{-1} = \begin{pmatrix}
		\Sigma_{11}^{-1} + \Sigma_{11}^{-1}\Sigma_{12}(\Sigma/\Sigma_{11})^{-1}\Sigma_{21}\Sigma_{11}^{-1} &-\Sigma_{11}^{-1} \Sigma_{12} (\Sigma/\Sigma_{11})^{-1} \\ 
		-(\Sigma/\Sigma_{11})^{-1}\Sigma_{21}\Sigma_{11}^{-1} & (\Sigma/\Sigma_{11})^{-1}
	\end{pmatrix}.
\end{equation}
This formula has the remarkable property that only the inverses 
\begin{equation}
	\Sigma_{11}^{-1} \text{\ and\ } (\Sigma/\Sigma_{11})^{-1}
\end{equation}
appear in it. With our choice of $W$ in equation (\ref{eqn.choice}) the Schur complement $M/E$ is block diagonal. This means that also its inverse is block diagonal. Indeed, we have
\begin{equation}
	(M/E)^{-1} = \begin{pmatrix}
		(M[\alpha]/E)^{-1} & 0 \\
		0 & (M[\beta]/E)^{-1}
	\end{pmatrix}.
\end{equation}
This allows us to write the inverse of $M$ explicitly:
\begin{equation}
	M^{-1} = \begin{pmatrix}
		(M[\alpha]/E)^{-1} & -(M[\alpha]/E)^{-1} D E^{-1} & 0 \\
		- E^{-1} D^T (M[\alpha]/E)^{-1} & \Xi & - E^{-1} F (M[\beta]/E)^{-1} \\
		0 & - (M[\beta]/E)^{-1} F^T E^{-1} & (M[\beta]/E)^{-1}
	\end{pmatrix},
\end{equation}
with
\begin{equation}
	\Xi = E^{-1} + E^{-1}\left(D^T(M[\alpha]/E)^{-1}D + F (M[\beta]/E)^{-1} F^T\right)E^{-1}.
\end{equation}
We see that the inverse of $M$ has zeroes in exactly those places where the undetermined matrix $W$ used to be.

@@

\subsection{The semidefinite case}
In the previous sections we have assumed that $E=M[\gamma]$ is non-singular so that $E^{-1}$ exists. In this section we will only assume that $E$ is positive semi-definite. It is in this case that our results differ from those in \cite{smith}. Because of this we give new proves for the following results that do not make a distinction between the definite and semi-definite case. We start by noting some properties of symmetric matrices that are also positive semi-definite.

\begin{lemma}\label{lem.symm}
	Let $\Sigma$ be a symmetric matrix that is partitioned as in equation (\ref{eqn.sigma}). If $\Sigma$ is positive semidefinite then we have
	\begin{enumerate}
		\item $\mathcal{C}(\Sigma_{12})\subset \mathcal{C}(\Sigma_{11})$ \\
		\item $\mathcal{R}(\Sigma_{21})\subset \mathcal{R}(\Sigma_{11})$ \\
		\item $r(\Sigma) = r(\Sigma_{11}) + r(\Sigma_{22})$
	\end{enumerate}
	Here $\mathcal{C}(\cdot)$ denotes the span of the column vectors of a matrix, $\mathcal{R}(\cdot)$ denotes the row space, and $r(\cdot)$ denotes the rank of a matrix.  
\end{lemma}

\begin{proof}
	The proof is given in appendix \ref{app.symm}. \qed
\end{proof}

In the following we no longer assume that the matrices are non-singular. It turns out that we can still define the Schur complement of a matrix. For a matrix $E$ we will denote by $E^-$ a generalized inverse of $E$. $E^-$ is a matrix that satisfies
\begin{equation}
	E E^{-}E = E.
\end{equation}
A particular example of a generalized inverse is the Moore-Penrose inverse $E^+$ \cite{roger}. For now, we will not assume that we are dealing with $E^+$ but keep the discussion general. Given a particular choice of generalized inverse we define
\begin{equation}
	\Sigma/\Sigma_{11} = \Sigma_{22} - \Sigma_{21}\Sigma_{11}^-\Sigma_{12}.
\end{equation}
If $\Sigma$ is a symmetric matrix that is also positive semi-definite it turns out that the Schur complement does not depend on the choice of the particular generalized inverse used in its definition.

\begin{lemma} Let $\Sigma$ be a symmetric matrix that is partitioned as in equation (\ref{eqn.sigma}). If $\Sigma$ is positive semidefinite then
\begin{equation}
	\Sigma/\Sigma_{11} = \Sigma_{22} - \Sigma_{21}\Sigma_{11}^-\Sigma_{12}
\end{equation}	
does not depend on the choice of the generalized inverse $\Sigma_{11}^-$.
\end{lemma}

\begin{proof}
	The proof is given in appendix \ref{app.symm}. \qed
\end{proof}

The next theorem characterizes positive symmetric matrices in terms of their Schur complement:

\begin{theorem}\label{theo.charact}
	Let $\Sigma$ be a symmetric matrix that is partitioned as in equation (\ref{eqn.sigma}) where $\Sigma_{11}$ and $\Sigma_{22}$ are square. Then $\Sigma\ge 0$ if and only if $\Sigma_{11}\ge 0$, $\mathcal{C}(\Sigma_{12})\subset \mathcal{C}(\Sigma_{11})$, and $\Sigma/\Sigma_{11}\ge 0$. 
\end{theorem}

\begin{proof}
	See \cite[Theorem 1.20]{horn} and \cite{albert}. \qed
\end{proof}

In equation (\ref{eqn.choice}) we defined $W$ using the inverse $E^{-1}$. We are now going to define $W$ using a general inverse instead. The next theorem shows that this again defines a positive semidefinite completion of $M$. The following theorems then show that the other properties also carry over to the semidefinite case.

\begin{theorem}
	Let $M$ be as in equation (\ref{eqn.corrmatrix}). Let both $M[\alpha]$ and $M[\beta]$ be positive semidefinite. Setting 
	\begin{align}
		W & = DE^-F \\
		  & = M[ \alpha-\gamma, \gamma]M[\gamma]^-M[ \gamma, \beta-\gamma]		
	\end{align}
	turns $M$ into a positive semidefinite matrix.
\end{theorem}

\begin{proof}
	Since both $M[\alpha]$ and $M[\beta]$ are positive semidefinite it follows from theorem \ref{theo.charact} that
	\begin{align}
		E & = M[\gamma] \ge 0 \\
		\mathcal{C}(D^T) & = \mathcal{C}(M[ \alpha-\gamma, \gamma]^T) \subset \mathcal{C}(E) \\
		\mathcal{C}(F) & = \mathcal{C}(M[ \gamma, \beta-\gamma ]) \subset \mathcal{C}(E) \\
		M[\alpha] / E & \ge 0 \\
		M[\beta] / E & \ge 0
	\end{align}
	From this it follows that
	\begin{align}
		E & = M[\gamma] \ge 0 \\
		\mathcal{C}(D^T\ F) & = \mathcal{C}( M[\gamma, I - \gamma] ) \subset  \mathcal{C}(E) \\
		M/E & = \begin{pmatrix}
			M[\alpha] / E & 0 \\
			0 & M[\beta] / E
		\end{pmatrix} \ge 0
	\end{align}
	Using theorem \ref{theo.charact} in the other direction now yields the result. \qed
\end{proof}

@@

\begin{thebibliography}{MM}
	\bibitem{heston} Steven L. Heston, A Closed-Form Solution for Options with Stochastic Volatility with Applications to Bond and Currency Options, The Review of Financial Studies, \textbf{6}(2), April 1993, 327--343.
	\bibitem{hayns} Emilie Virginia Haynsworth, Determination of the inertia of a partitioned Hermitian matrix,  Linear  Algebra and its Applications \textbf{1}, 1968, 73--81. 
	\bibitem{horn} Roger A. Horn, Basic  Properties  of  the  Schur  Complement. In Fuzhen Zhang (Ed.) The Schur complement and its application (17--46). Springer, 2005.
	\bibitem{rao} C. Radharkishna Rao, Linear Statistical Inference and its Applications, 2nd edition, John Wiley \& Sons, Inc, 2002.
	\bibitem{cottle} Richard W. Cottle, Manifestations of the Schur complement, Linear Algebra and its Applications \textbf{8}, 1974, 189--211. 
	\bibitem{smith} Smith, R. (2008). The positive definite completion problem revisited Linear Algebra and its Applications  429(7), 1442-1452. https://dx.doi.org/10.1016/j.laa.2008.04.020
	\bibitem{ouellette} Ouellette, D. (1981). Schur complements and statistics Linear Algebra and its Applications  36, 187-295. https://dx.doi.org/10.1016/0024-3795(81)90232-9
	\bibitem{roger} Penrose, R. (1955). A generalized inverse for matrices Mathematical Proceedings of the Cambridge Philosophical Society  51(3), 406-413. https://dx.doi.org/10.1017/s0305004100030401
	\bibitem{albert} Albert, A. (1969). Conditions for Positive and Nonnegative Definiteness in Terms of Pseudoinverses SIAM Journal on Applied Mathematics  17(2), 434-440. https://dx.doi.org/10.1137/0117041
\end{thebibliography}

\appendix

\section{Symmetric matrices \& the Schur complement}
\label{app.symm}
